---
title: "Working with chromosomes"
vignette: >
  %\VignetteIndexEntry{Working with chromosomes}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup
#| warning: FALSE
#| message: FALSE
library(chroordinate)
library(readr)
library(readxl)
library(dplyr)
library(ggplot2)

```

Chromosome names can be difficult to work with for the following reasons:

1. They are a mixture of numbers (1-22) and letters (X and Y). This can make
automatic assigning of data types difficult.

2. They are ordered - it is traditional to present chromosomes in the order
1-22 followed by the X and Y chromosomes.

3. They can be inconsistently formatted. For example the "chromosome" column
of a table may contain information written as "1", 1, "chr1", "1p" etc.

`chroordinate` is designed to assist with some of these issues.

## Reading chromosome names from Excel files

Here is an example of a small Excel file containing a column of chromosome information
for autosomes 1-6. We can read it into R using `read_excel`.

```{r}

chr_excel_short_autosomes <- read_excel(path = "data/chr_short_autosomes.xlsx",
                              col_types = "guess")

knitr::kable(chr_excel_short_autosomes)

```

As col_types is set to "guess", `read_excel` notes that there are only numbers
in the "chromosome" column and so assigns the "numeric" data type.

```{r}

typeof(chr_excel_short_autosomes$chromosome)

unique(chr_excel_short_autosomes$chromosome)

```

If the Excel file also contains information on the X and Y chromosomes, then
`read_excel` notes that there are 
both numbers (autosomes) and letters (sex
chromosomes) in the column and so chooses type "character".

```{r}

chr_excel_short_all <- read_excel(path = "data/chr_short_all.xlsx",
                              col_types = "guess")

knitr::kable(chr_excel_short_all)

```

```{r}

typeof(chr_excel_short_all$chromosome)

unique(chr_excel_short_all$chromosome)

```

```{r}
#| eval: FALSE
#| include: FALSE

# Code for generating long dataframe

autosome_df <- data.frame(
  "chromosome" = c(rep(seq(from = 1, to = 22, by = 1), 
                       times = 100))) |> 
  arrange(chromosome) 

sex_chrom_df <- data.frame(
  "chromosome" = c(rep("X", times = 100),
                   rep("Y", times = 100)))

chr_df_long <- rbind(autosome_df, sex_chrom_df) |> 
  mutate(region = row_number(),
         coverage = round(runif(n = 2400, min = 800, max = 1000), 0))

```

This behaviour then changes if the Excel is very long, which is common for
large genetic datasets from next generation sequencing.
In this examples, autosomes are listed first in 
the Excel and rows for sex chromosomes only appear after ~2000 rows.
This is a problem, because the "guess" argument checks only the first 1000 rows,
and decides that "chromosome" will be data type numeric.
The sex chromosomes are then converted into NA, with warnings given
(suppressed in this vignette).

```{r}
#| warning: FALSE

chr_excel_long_guess <- read_excel(path = "data/chr_long.xlsx",
                              col_types = "guess")

typeof(chr_excel_long_guess$chromosome)

unique(chr_excel_long_guess$chromosome)

```

One option is to specify that the chromosome column is type "text" when
reading the file in.

```{r}

chr_excel_long_text <- read_excel(path = "data/chr_long.xlsx",
                              col_types = c("text", "numeric", "numeric"))

typeof(chr_excel_long_text$chromosome)

unique(chr_excel_long_text$chromosome)

```

This solves the problem of the missing sex chromosomes.
But it introduces a new problem that the chromosomes are not ordered correctly
when plotted: chromosome 10 is ordered after chromosome 1.

```{r}

ggplot(chr_excel_long_text, aes(x = chromosome, y = coverage)) +
  geom_boxplot() +
  theme_bw()

```

`chroordinate` can be used to solve this issue by specifying the col_type as
"text" and then converting the chromosome column into a factor using
`factorise_chromosome`.

```{r}

chr_excel_long_factor <- read_excel(path = "data/chr_long.xlsx",
                              col_types = c("text", "numeric", "numeric")) |> 
  mutate(chromosome = factorise_chromosome(chromosome))

typeof(chr_excel_long_factor$chromosome)

str(chr_excel_long_factor$chromosome)


```

This means that the chromosomes are in the correct ordered when plotted.

```{r}

ggplot(chr_excel_long_factor, aes(x = chromosome, y = coverage)) +
  geom_boxplot() +
  theme_bw()

```

## Reading chromosome names from CSV files

Unlike `readxl`, the `guess_max` argument in `read_csv` checks 1000 rows spread
throughout the file, not just the first 1000 rows.
This means that sex chromosomes are not converted into NAs, even if the file
is long and col_types are guessed.

```{r}

chr_csv_long <- read_csv(file = "data/chr_long.csv",
                         col_types = "?")

typeof(chr_csv_long$chromosome)

unique(chr_csv_long$chromosome)

```

However this still leaves the issue of chromosome ordering,
which can be solved by supplying the correct order as `levels_chromosomes`
within the col_types argument.

```{r}

chr_csv_long_factor <- read_csv(file = "data/chr_long.csv",
                         col_types = list(
                           "chromosome" = col_factor(levels = levels_chromosomes()$all)
                         ))

ggplot(chr_csv_long_factor, aes(x = chromosome, y = coverage)) +
  geom_boxplot() +
  theme_bw()

```
